\documentclass[11pt,a4paper]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\graphicspath{{images/}}
\usepackage{booktabs}
\usepackage{array}
\usepackage{geometry}
\usepackage{setspace}
\newcommand{\degree}{$^\circ$}
\usepackage{textcomp}
\usepackage{threeparttable}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{float}
\usepackage{caption}
\usepackage{longtable}

% Unicode character support
\usepackage{newunicodechar}
\newunicodechar{⁵}{\textsuperscript{5}}
\newunicodechar{⁴}{\textsuperscript{4}}
\newunicodechar{⁰}{\textsuperscript{0}}
\newunicodechar{ⁿ}{\textsuperscript{n}}
\newunicodechar{≤}{\leq}
\newunicodechar{≈}{\approx}
\newunicodechar{−}{\ensuremath{-}}
\newunicodechar{₀}{$_0$}
\newunicodechar{₁}{$_1$}
\newunicodechar{₂}{$_2$}
\newunicodechar{₃}{$_3$}
\newunicodechar{₄}{$_4$}
\newunicodechar{₅}{$_5$}
\newunicodechar{₆}{$_6$}
\newunicodechar{₇}{$_7$}
\newunicodechar{₈}{$_8$}
\newunicodechar{₉}{$_9$}
\newunicodechar{⁷}{\textsuperscript{7}}
\newunicodechar{√}{\sqrt{}}
\newunicodechar{ⱼ}{\textsubscript{j}}
\newunicodechar{ₖ}{_k}
\newunicodechar{≥}{\geq{}}
\newunicodechar{⁶}{\textsuperscript{6}}
\newunicodechar{⁻}{\textsuperscript{-}}

% Language
\usepackage[english]{babel}

% Page setup
\geometry{margin=1in}
\onehalfspacing

% Hyperref setup
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}

% Custom commands
\newcommand{\phisym}{\varphi}
\newcommand{\fzero}{\ensuremath{f_0}}
\newcommand{\phiband}[1]{$\phisym$-band_{#1}}

% Bibliography
\usepackage[style=apa,backend=biber,sorting=nyt]{biblatex}
\addbibresource{spectral_differentiation_refs.bib}

\begin{document}

% Title
\begin{center}
{\LARGE \textbf{Spectral Differentiation:}}\\[0.1cm]
{\LARGE \textbf{Within-Band Oscillatory Organization as a Cognitive}}\\[0.1cm]
{\LARGE \textbf{and Developmental EEG Biomarker}}\\[0.3cm]
{\Large \textbf{A Coordinate System for Human Oscillation Bands}}\\[1cm]

{\large Michael Lacy}\\[0.3cm]
{\small Independent Researcher}\\[0.3cm]
{\small michael.lacy@gmail.com}\\[0.5cm]

\today
\end{center}

\vspace{0.5cm}

%==============================================================================
% ABSTRACT
%==============================================================================
\begin{abstract}
Since Berger's discovery of the alpha rhythm in 1929, EEG frequency bands have been defined by clinical convention rather than principled theory. Although log-frequency spacing has been observed \parencite{penttonen2003natural} and the golden ratio ($\phisym \approx 1.618$) has been proposed as the inter-band ratio \parencite{pletzer2010golden,kramer2023golden}, no study has formally tested these claims against alternatives on large-scale data. Here we present the first model comparison of EEG frequency scaling using 4,572,489 FOOOF-detected oscillatory peaks from 9 datasets spanning 2,097 subjects aged 5--70. Log-frequency description is decisively superior to linear ($\Delta$BIC $>$ 50). Band boundaries are genuine spectral features, not 1/f artifacts ($p < 0.0001$ vs.\ aperiodic null). Among seven candidate ratios, $\phisym$ is the best fixed-ratio model for boundary positions (BIC $= -24.7$); subject-level bootstrap resampling ($N = 2{,}097$ subjects, stratified by dataset) localized all five density troughs in 100\% of 1,000 iterations and yielded a 95\% CI of [1.609, 1.623] for the global inter-trough geometric mean ratio, containing $\phisym = 1.618$ ($p = 0.90$) while excluding all other named alternatives tested ($\sqrt{2}$, $e-1$, $e$, octave; all $p < 0.0001$). A $\phisym$-lattice coordinate system produces the simplest within-band enrichment profiles ($R^2 = 0.968$, with boundaries within 2\% of locally optimal positions), but within bands, no fine-structure exists at $\phisym$-specific positions---enrichment profiles are smooth ramps (theta, beta, gamma) or a single mountain (alpha) with no preferred landmarks. We introduce \emph{spectral differentiation}---the degree to which oscillatory peaks concentrate at specific within-band positions---as a novel class of EEG biomarker. The core cognitive finding is that beta-low spectral differentiation predicts reasoning ability in a single adult sample ($\rho = -0.27$, $N = 203$, surviving FDR correction, age partialing, and IRASA replication); the same sample yields a broader but hypothesis-generating cognitive map (31 FDR survivors across 4 tests and 4 bands, requiring independent replication). Spectral differentiation tracks a cross-sectional inverted-U trajectory from childhood through aging (60 + 41 FDR-significant age correlations across two independent datasets, with the caveat that the vertex at $\sim$33--35 years is driven by a cross-cohort splice rather than a within-cohort turning point), dissociates externalizing from internalizing psychopathology (18 + 7 FDR survivors), is uninformative about personality (0/11,970 FDR), and is stable as an individual trait over 5 years (ICC $= 0.75$). The strongest enrichment feature, the core cognitive association, and the developmental trajectory replicate under IRASA, a non-parametric alternative to FOOOF; the broader cognitive map does not, likely attributable in substantial part to IRASA's lower per-subject peak yield in the bands carrying the strongest cognitive signal. We distinguish robust from provisional findings. Log-frequency scaling, the five-trough structure, and the near-optimality of $\phisym$-lattice boundaries are robust across datasets, methods, and analytical choices. The biomarker properties of spectral differentiation---developmental trajectory, reliability, and the strongest cognitive and psychopathology associations---replicate across extraction methods (FOOOF and IRASA), though the breadth of the cognitive map (31 FDR survivors) relies on a single adult dataset ($N = 203$) and requires independent replication. A post hoc inhibitory-circuit interpretation, in which the $\alpha$/$\beta$ trough's developmental deepening tracks parvalbumin-positive interneuron maturation and its depth dissociates externalising from internalising psychopathology, generates falsifiable pharmacological predictions but awaits direct neurochemical testing. The coordinate system is the study's methodological contribution; the biomarker it reveals---spectral differentiation---is its empirical contribution.

\vspace{0.3cm}
\textbf{Keywords:} EEG, frequency bands, logarithmic scaling, golden ratio, spectral parameterization, oscillatory architecture, biomarker, cognitive development, spectral differentiation
\end{abstract}

\paragraph{Impact statement.} Analysis of 4.6 million brain oscillation peaks reveals that EEG frequency bands follow a logarithmic organization, and that the pattern of oscillatory peaks within each band is a promising biomarker of brain maturation and selected cognitive differences.

\section*{Significance Statement}
Brain oscillation bands---theta, alpha, beta, gamma---are among the most widely used constructs in neuroscience, yet their boundaries have never been formally optimized. We show that optimizing band boundaries on 4.6 million EEG oscillation peaks reveals a new measurable phenotype: \emph{spectral differentiation}, the shape of the within-band peak distribution. A logarithmic coordinate system with boundaries near the golden ratio $\phisym = 1.618$ (subject-level bootstrap 95\% CI: 1.609--1.623, excluding all other named constants tested) provides the measurement-optimal framework for this phenotype---best for profile simplicity, with boundaries within 2\% of locally optimal positions. Spectral differentiation is modestly associated with reasoning ability (peak $|\rho| = 0.27$; $\sim$2\% unique variance after age; single dataset, $N = 203$, requiring replication), tracks brain maturation from childhood through aging (cross-sectional; vertex estimate requires longitudinal confirmation), distinguishes psychiatric phenotypes, is uninformative about personality and medical variables (0 FDR survivors across $>$14{,}000 tests), and is stable over 5 years (ICC $= 0.75$), yet captures information distinct from spectral power, peak frequency, or aperiodic slope. The strongest enrichment feature, the core cognitive result, and the developmental trajectory replicate under IRASA, a non-parametric alternative; weaker features do not, and we report which is which. The developmental trajectory of the $\alpha$/$\beta$ band boundary---immature in childhood, deepening through adolescence on a timeline consistent with parvalbumin-positive interneuron maturation---suggests that spectral differentiation may index inhibitory circuit integrity, a post hoc interpretation that generates specific pharmacological predictions. The coordinate system is the study's methodological contribution; the biomarker it reveals is its empirical contribution.

\newpage
\tableofcontents
\newpage

%==============================================================================
% INTRODUCTION
%==============================================================================
\input{spectral_differentiation_introduction}

%==============================================================================
% RESULTS
%==============================================================================
\section{Results}

\input{spectral_differentiation_results_part1}

\input{spectral_differentiation_results_part2}

\input{spectral_differentiation_results_part3_4}

%==============================================================================
% DISCUSSION
%==============================================================================
\input{spectral_differentiation_discussion}

%==============================================================================
% METHODS
%==============================================================================
\input{spectral_differentiation_methods}

%==============================================================================
% DATA AVAILABILITY
%==============================================================================
\section*{Data and Code Availability}

Extracted spectral peaks (14.7 million FOOOF and 11.7 million IRASA peaks), per-subject enrichment profiles, subject demographics, and all statistical analysis outputs are available on Dryad (\url{https://doi.org/10.5061/dryad.1vhhmgr8t}). All analysis code is available at \url{https://github.com/neurokinetikz/research} under the MIT License. Raw EEG data are publicly available from their original sources: EEGMMIDB (PhysioNet, \url{https://physionet.org/content/eegmmidb/}; \cite{goldberger2000physiobank}), LEMON (Max Planck Institute, \url{http://fcon_1000.projects.nitrc.org/indi/retro/MPI_LEMON.html}; \cite{babayan2019lemon}), HBN (Child Mind Institute, \url{http://fcon_1000.projects.nitrc.org/indi/cmi_healthy_brain_network/}; \cite{alexander2017hbn}), Dortmund Vital Study (OpenNeuro, \url{https://doi.org/10.18112/openneuro.ds005385.v1.0.3}), and CHBMP (CAN-BIND data sharing portal). CHBMP-derived data are not included in the Dryad package due to its CC BY-NC-SA license but can be reproduced from the source data using the provided extraction code. LEMON cognitive test scores and HBN CBCL psychopathology scores are available from their respective data portals under the original data use agreements.

%==============================================================================
% ACKNOWLEDGMENTS
%==============================================================================
\section*{Acknowledgments}

The author thanks the teams behind the publicly available datasets that made this work possible: the PhysioNet EEGMMIDB \parencite{goldberger2000physiobank}, the MPI Leipzig Mind-Brain-Body LEMON study \parencite{babayan2019lemon}, the Child Mind Institute Healthy Brain Network \parencite{alexander2017hbn}, the Dortmund Vital Study, and the Canadian Biomarker Integration Network for Depression (CAN-BIND) CHBMP. Computational resources were provided by Google Cloud Platform.

%==============================================================================
% AUTHOR CONTRIBUTIONS
%==============================================================================
\section*{Author Contributions}

Michael Lacy: Conceptualization, Data curation, Formal analysis, Investigation, Methodology, Software, Validation, Visualization, Writing -- original draft, Writing -- review \& editing.

%==============================================================================
% COMPETING INTERESTS
%==============================================================================
\section*{Competing Interests}

The author declares no competing interests.

%==============================================================================
% FUNDING
%==============================================================================
\section*{Funding}

No external funding was received for this work.

%==============================================================================
% REFERENCES
%==============================================================================
\newpage
\printbibliography

\end{document}
